
---
title: "Reproducible Research - PA1 - Activity Monitoring Data - MD"
output: html_document
---

##Loading and preprocessing the data  
1. **Loading the data**

```{r}
rdata <- read.csv("activity.csv",stringsAsFactors=FALSE)

```
2. **Formatting the data to omit na and factor date**

```{r}
adata <- na.omit(rdata)
adata$date <- as.factor(as.character(adata$date))
```

##Total number of steps taken per day
1.**Create data for total number of Steps Taken**
```{r}
dqa <- tapply(adata$steps, adata$date, sum, na.rm = TRUE)
dq1a <- cbind.data.frame(date = names(dqa),totalsteps = unname(dqa))

```

2. **Plot for Frequency of Steps**
```{r, echo=FALSE}
hist(dq1a[,2], xlab = "Total Steps Per Day", ylab = "Number of Days", main = "Frequency of Total Steps in a day", col = "green")  
```  

3.**Calculate Mean and Median**
```{r}
meansteps <- mean(dq1a$totalsteps, na.rm = TRUE)
mediansteps <- median(dq1a$totalsteps, na.rm = TRUE)
print(sprintf("Mean total steps taken per day: %f ", meansteps))
print(sprintf("Median total steps taken per day: %f ", mediansteps))
```  


##Average daily activity pattern
1. **Getting Data for Time Series**
```{r}
dqa <- tapply(adata$steps, adata$interval, mean,na.rm = TRUE)
dq2a <- cbind.data.frame(interval = as.integer(names(dqa)),avgsteps = unname(dqa))
```  

**Time Series plot**
 
```{r, echo=FALSE}
plot(dq2a$interval, dq2a$avgsteps, type = "l", xlab = "5-min Interval", ylab = "Average Num of Steps", main = "Average Daily Activity Pattern")
```   

2. **Maximum number of steps - for 5 min interval range**
```{r}
maxavginterval <- dq2a$interval[dq2a$avgsteps == max(dq2a$avgsteps)]
print(sprintf("5-min time Interval with maximum average steps taken per day: %i ", maxavginterval)) 
``` 

##Imputing missing values
1. **Total Number of Missing Values**
```{r}
nacount <- nrow(rdata[!complete.cases(rdata),])
print(sprintf("Row count with NA data: %i ", nacount))
``` 

2.**Filling in missing value Strategy**
```{r}
ndata <- rdata 
sapply(unique(rdata$interval), 
      function(x) 
      ndata[!complete.cases(ndata) & (ndata$interval == x),1] <<- dq2a$avgsteps[dq2a$interval == x])
```  
3.**New Data Set with filled value**
```{r}
dqc <- tapply(ndata$steps, ndata$date, sum, na.rm = TRUE)
dq1c <- cbind.data.frame(date = names(dqc),totalsteps = unname(dqc))

```  

4. * New Plot with filled value*  
```{r, echo=FALSE}

hist(dq1c[,2], xlab = "Total Steps Per Day", ylab = "Number of Days", main = "Frequency of Total Steps in a day", col = "blue") 

``` 
**New mean and median calculation with filled value**
```{r}
meansteps <- mean(dq1c$totalsteps) 
mediansteps <- median(dq1c$totalsteps)
print(sprintf("Mean total steps taken per day: %f ", meansteps))
print(sprintf("Median total steps taken per day: %f ", mediansteps))

```  


##Differences in activity pattersn weekdays and weekends  
1. **Create new dataset**
```{r}
wd <- weekdays(as.Date(ndata$date))
ndata$dayTypes <- factor(wd, levalns <- c('weekday', 'weekend'))
ndata$dayTypes[] <- 'weekday'
ndata$dayTypes[wd %in% c('saturday', 'Sunday')] <- 'weekend'

```{r, echo=FALSE}
library (plyr)
library (lattice)
```  
```{r}

dq3a <- ddply(ndata,
             .(interval, dayTypes),
             summarise,
             avgSteps = mean(steps, na.rm=TRUE))


```  

2. **Comparison Plot**  

```{r, echo=FALSE}
xyplot(avgSteps ~ interval | dayTypes, 
       data = dq3a,
       type = 'l',
       layout = c(1,2),
       ylab = "Average number of steps")

``` 


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
